Practical Optional Types for Clojure

### 2015-02-28 formal-model.tex
3.4

    {\TAssoc}
  \end{mathpar}
  \begin{mathpar}
    {\BAssoc}
    {\BGet}
    {\BGetMissing}
  \end{mathpar}
  \begin{mathpar}
    \HMapsubtyping{}
  \end{mathpar}
  \caption{HMap Terms, Types, Semantics, Subtyping and Typing Rules}
  \label{main:figure:hmapsyntax}
\end{figure}

@@

\begin{figure}[h]
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \mininonreflectiveexpsyntax{}
    \\

@@

    \tatypesyntax{}\\
    \typehintenvsyntax{}\\
    \classtableallsyntax{}
  \end{altgrammar}
  $$
 \classtablelookupsyntax{}
 \begin{mathpar}
  \begin{altgrammar}
    \convertjavatypenil{}
  \end{altgrammar}
  \begin{altgrammar}
    \convertjavatypenonnil{}
  \end{altgrammar}
\end{mathpar}
  \begin{mathpar}
    {\TNewStatic}

@@

    %{\TInstance}
  \end{mathpar}
  \begin{mathpar}
    \BField{}

@@

%    \RLetHint{}
  \end{mathpar}
  \caption{Java Interoperability Syntax, Semantics, Typing Rules and Rewrite Relation}
  \label{main:figure:javatyping}
\end{figure}

@@

  $$
  \begin{altgrammar}
    \e{} &::=& \ldots   \localhintsyntax{} \alt \lethintsyntax{} &\mbox{Expressions}\\
            &\alt& \reflectiveexpsyntax{} 
  \end{altgrammar}
  $$

We extend the syntax with type hinted expressions, Java field lookups and calls to
methods and constructors. We model the syntax after the `dot' special
form to prevent ambiguity---\clj{(.fld e)} is now \fieldexp{\fld{}}{\e{}},
\clj{(.mth e es*)} is $\methodexp{\mth{}}{\e{}}{\overrightarrow{es}}$
and \clj{(.class es*)} is $\newexp{\class{}}{\overrightarrow{es}}$.
The Java expressions come without typing rules because Typed Clojure
only reasons about resolved reflection
(as demonstrated in \secref{sec:overviewjavainterop}).

Now we model the compiler's reflection-resolution algorithm.
\figref{main:figure:javatyping} gives the syntax for non-reflective Java calls.
The method ${\methodstaticexp {\classhint{1}}
                             {\overrightarrow {\classhint{i}}}
                             {\classhint{2}}
                             {\mth{}} {\e{}} {\overrightarrow{\e{i}}}}$
is a non-reflective call to the \mth{} method on class {\class{1}}, 
with Java signature 
${\classhint{2}}\ \mth{}\ (\overrightarrow {\classhint{i}});$.
The field {\fieldstaticexp {\classhint{1}} {\classhint{2}} {\fld{}} {\e{}}}
calls the field on class {\classhint{1}} with Java signature
${\classhint{2}}\ \fld{};$.
The constructor {\newstaticexp {\overrightarrow{\classhint{i}}} {\classhint{}} 
                               {\class{}} {\overrightarrow{\e{i}}}}
calls the constructor with Java signature
${\classhint{}}\ ({\overrightarrow{\classhint{i}}});$.

@@

To obtain non-reflective calls we define our rewrite relation
\rewrite {\taenv{}} {\e{}} {\ep{}}
which rewrites \e{} to a possibly-less reflective expression
\ep{} with respect to type hint environment
\taenv{} and Java class table \ct{}.

For example,
R-FieldElimRefl emits a non-reflective field if it can
find a field matching the type hint inferred on \ep{}.
  $$
    \RFieldElimRefl{}
  $$

Type hint inference is modelled by
\tajudgement {\taenv{}} {\hastype {\e{}} {\tatype{}}}
which infers the (possibly-unknown) type hint \tatype{} of expression \e{} in environment \taenv{}.
%$$
%\TALetHint{}
%$$

For demonstration purposes let us rewrite
a simple field expression, with the assumptions that
that we have a class \Point{} with a field \getx{} of Java type \Number
and \intaenv {\taenv{}} {\x{}} {\Point{}}.
In rewriting
the expression \fieldexp{\getx}{\x{}},
{\Point{}} is inferred as the type hint of \x{}
and \Number{} is the field type
by \fieldtypeliteral{} (\figref{main:figure:javatyping}).
Now we just plug in the new information into our new expression
$$
\fieldstaticexp{\Point}{\Number}{\getx}{\x{}}.
$$

@@

To continue our example, let us assume \propenv{} = {\isprop{\Point}{\x{}}}.
T-FieldStatic checks \x{} under the non-nilable type {\Point}
and returns the nilable type {\Union{\Nil}{\Number}}.

The other rules T-MethodStatic and T-NewStatic work similarly (\figref{main:figure:javatyping}), varying
in the choice of nilability in the conversion function---method returns
are nilable and constructor returns are non-nilable.

The operational semantics B-Field, B-New and B-Method (\figref{main:figure:javatyping}) simply evaluate their
arguments and call the relevant JVM operation---\secref{sec:metatheory}
states our exact assumptions about each.

@@

 \disptablesyntax{} \\
\end{altgrammar}
$$
  \begin{mathpar}
    \Multisubtyping{}
  \end{mathpar}
  \begin{mathpar}
    \isapropsfigure{}
  \end{mathpar}
  \begin{mathpar}
    \TDefMulti{}

    \TDefMethod{}
  \end{mathpar}
  \getmethodfigure{}
  \begin{mathpar}
    \BDefMethod{}
    \BDefMulti{}
    \BBetaMulti{}
  \end{mathpar}
\caption{Multimethod Syntax, Subtyping, Typing Rules and Semantics}
\label{main:figure:mmsyntax}
\end{figure}

To ease presentation, we present \emph{immutable}
multimethods (\figref{main:figure:mmsyntax}). \defmethodliteral{} returns a new extended multimethod
without changing the original multimethod. \egref{example:rep} is now written
\begin{minted}{clojure}
(let [path (defmulti [Any -> (U nil String)] class)]
  (let [path (defmethod path String [x] x)]
    (let [path (defmethod path File [^File x] 
                 (.getPath x))]
      (path "dir/a")))) ;=> "a"
\end{minted}
where subsequent bindings in the binder is sugar for nested let expressions.

>>>>

    {\TAssoc}
  \end{mathpar}
  \begin{mathpar}
    \HMapsubtyping{}
  \end{mathpar}
  \begin{mathpar}
    {\BAssoc}
    {\BGet}
    {\BGetMissing}
  \end{mathpar}
  \caption{HMap Syntax, Typing and Operational Semantics}
  \label{main:figure:hmapsyntax}
\end{figure}

@@

\begin{figure}[h]
  \footnotesize

  $$
  \begin{altgrammar}
  \end{altgrammar}
  $$
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots   \localhintsyntax{} \alt \lethintsyntax{} &\mbox{Expressions}\\
            &\alt& \reflectiveexpsyntax{} \\
     &\alt& \mininonreflectiveexpsyntax{}
    \\

@@

    \tatypesyntax{}\\
    \typehintenvsyntax{}\\
    \classtableallsyntax{}
  \end{altgrammar}
  $$
  \begin{mathpar}
    {\TNewStatic}

@@

    %{\TInstance}
  \end{mathpar}
 \classtablelookupsyntax{}
 \begin{mathpar}
  \begin{altgrammar}
    \convertjavatypenil{}
  \end{altgrammar}
  \begin{altgrammar}
    \convertjavatypenonnil{}
  \end{altgrammar}
\end{mathpar}
  \begin{mathpar}
    \BField{}

@@

%    \RLetHint{}
  \end{mathpar}
  \caption{Java Interoperability Syntax, Typing and Operational Semantics}
  \label{main:figure:javatyping}
\end{figure}

@@

We extend the syntax in \figref{main:figure:javatyping} with
type hinted expressions, reflective and non-reflective Java field lookups and calls to
methods and constructors. 
We model the syntax after the `dot' special
form to prevent ambiguity---\clj{(.fld e)} is now \fieldexp{\fld{}}{\e{}},
\clj{(.mth e es*)} is $\methodexp{\mth{}}{\e{}}{\overrightarrow{es}}$
and \clj{(.class es*)} is $\newexp{\class{}}{\overrightarrow{es}}$.
The reflective expressions come without typing rules because Typed Clojure
only reasons about resolved reflection
(as demonstrated in \secref{sec:overviewjavainterop}).
%
The method call ${\methodstaticexp {\classhint{1}}
                             {\overrightarrow {\classhint{i}}}
                             {\classhint{2}}
                             {\mth{}} {\e{}} {\overrightarrow{\e{i}}}}$
is a non-reflective call to the \mth{} method on class {\class{1}}, 
with Java signature 
${\classhint{2}}\ \mth{}\ (\overrightarrow {\classhint{i}});$.
The field access {\fieldstaticexp {\classhint{1}} {\classhint{2}} {\fld{}} {\e{}}}
calls the field on class {\classhint{1}} with Java signature
${\classhint{2}}\ \fld{};$.
The constructor invocation {\newstaticexp {\overrightarrow{\classhint{i}}} {\classhint{}} 
                               {\class{}} {\overrightarrow{\e{i}}}}
calls the constructor with Java signature
${\classhint{}}\ ({\overrightarrow{\classhint{i}}});$.

@@

We model Clojure's reflection resolution algorithm as a 
rewrite relation
\rewrite {\taenv{}} {\e{}} {\ep{}}
which rewrites \e{} to a possibly-less reflective expression
\ep{} with respect to type hint environment
\taenv{} and Java class table \ct{}.
%
For example,
R-FieldElimRefl emits a non-reflective field if it can
find a field matching the type hint inferred on \ep{}.
  $$
    \RFieldElimRefl{}
  $$

Type hint inference is given by the judgement
\tajudgement {\taenv{}} {\hastype {\e{}} {\tatype{}}}
which infers the (possibly-unknown) type hint \tatype{} of expression \e{} in type hint environment \taenv{}.
We defer the remainder of the rules for rewriting and the definition of type hint inference
to the supplemental material.
%$$
%\TALetHint{}
%$$

As an example, we rewrite
a simple field reference, with the assumptions that
that the class \Point{} has a field $x$ of Java type \Number
and \intaenv {\taenv{}} {p} {\Point{}}.
In rewriting
the expression \fieldexp{x}{p},
{\Point{}} is inferred as the type hint of $p$
and \Number{} is the field type
by \fieldtypeliteral{} (\figref{main:figure:javatyping}).
Now we just plug in the new information into our new expression
$$
\fieldstaticexp{\Point}{\Number}{x}{p}.
$$

@@

To continue our example, assume \propenv{} = {\isprop{\Point}{p}}.
T-FieldStatic therefore produces the type
{\Union{\Nil}{\Number}}
for the entire expression.

The rules T-MethodStatic and T-NewStatic work similarly (\figref{main:figure:javatyping}), varying
in the choice of nilability in the conversion function---methods
can return nil but constructors cannot.

The evaluation rules B-Field, B-New and B-Method (\figref{main:figure:javatyping}) simply evaluate their
arguments and call the relevant JVM operation, which do not model---\secref{sec:metatheory}
states our exact assumptions about each.

@@

 \disptablesyntax{} \\
\end{altgrammar}
$$
  \begin{mathpar}
    \TDefMulti{}

    \TDefMethod{}
  \end{mathpar}
  \begin{mathpar}
    \isapropsfigure{}
  \end{mathpar}
  \begin{mathpar}
    \Multisubtyping{}
  \end{mathpar}
  \begin{mathpar}
    \BDefMethod{}
    \BDefMulti{}
    \BBetaMulti{}
  \end{mathpar}
  \getmethodfigure{}
\caption{Multimethod Syntax, Typing and Operational Semantics}
\label{main:figure:mmsyntax}
\end{figure}

To ease presentation, we present \emph{immutable}
multimethods (\figref{main:figure:mmsyntax}). \defmethodliteral{} returns a new extended multimethod
without changing the original multimethod. \egref{example:rep} is now written
\begin{minted}{clojure}
(let [path (defmulti [Any -> (U nil String)] class)]
  (let [path (defmethod path String [x] x)]
    (let [path (defmethod path File [^File x] 
                 (.getPath x))]
      (path "dir/a")))) ;=> "a"
\end{minted}

### 2015-02-28 metatheory.tex
3.4

Our proof for type soundness is similar to~\citet{TF10}. We add
errors and \wrong{} value and prove well-typed programs
rule out wrong programs.

@@

We use an extra lemma to support our main soundness lemma. Consistency
ensures that occurrence typing does not claim the outer 
\x{} is a keyword.

\begin{minted}{clojure}
(let [x 1]
  ((let [x :a]
     (fn [] x))))
\end{minted}

>>>>

Our proof for type soundness is similar to~\citet{TF10}. We add
errors and \wrong{} value and prove
well-typed programs do not go wrong.

@@

We use an extra lemma to support our main soundness lemma. Consistency
ensures that occurrence typing does not refer to variables
hidden inside a closure.
